Modelling BCG vaccination in the UK

### 2019-12-11 rmds/08-model-development.Rmd
complete dodd typos

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have indirect impacts, via onwards transmission, not captured in this estimate. Both of these limitations can be overcome using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes, via model simulation.

This chapter presents the development and parameterisation of a, semi-stochastic, dynamic model of TB transmission, incorporating BCG vaccination, in the UK born population of England. The key features of TB transmission, and BCG vaccination, are discussed with details of pertinent TB models given. An appropriate model structure for answering the study question is then outlined, along with a justification of the choices made and details of required sensitivity analyses. The model structure is then defined mathematically and parameterised using literature sources as well as data from the Enhanced TB Surveillance System (ETS), Labour Force Survey (LFS) and Office for National Statistics (ONS) (see Chapter \@ref(data)). The assumptions made during model building and parameterisation are highlighted in preparation for evaluation during model fitting (Chapter \@ref(model-fitting)).

@@

1. **Latency** - after an initial infection 5-10% of individuals develop symptomatic TB within 1-2 years. The majority of individuals enter a latent state in which they passively carry TB mycobacterium but are not symptomatic. Reactivation of the bacilli can then occur many years later due to a lose of immune control.[@Gideon2011a] Simplistically latent TB may be modeled with a single latent compartment[@Brooks-Pollock2010a], more commonly an additional transition rate between the susceptible and active disease states is added.[@Menzies2018] This represents rapid progression to active disease, and slower progression via a low risk latent stage. Both of these model structures have been shown to not fit activation data well.[@Menzies2018; @Ragonnet2017] More complex structures that are commonly used incorporate either parallel or serial latency (Figure \@ref(fig:latency-flow-diag)). Both of these structures incorporate both slow and fast latent periods and have been shown to produce identical activation dynamics.[@Ragonnet2017] This is unfortunate as they represent two disparate biological mechanisms, with the serial assumption representing decreasing risk over time for individuals and the parallel assumption suggesting that a subset of individuals are at a greater risk of developing active TB disease. For models that seek to investigate interventions targeted at latent cases this structural uncertainty is problematic. However, as BCG vaccination occurs prior to infection both structures will produce comparable results for study questions evaluating this intervention. The model presented here uses a serial latent structure. This is commonly used in the literature; simplifies modelling other aspects of TB; and has a plausible biological underpinning.[@Ragonnet2017]

@@

1. **Protection from active disease** - the BCG vaccine has been shown to primarily protect against the progression from latent to active TB disease (Chapter \@ref(background)). It has been shown to be highly protective in children[@Rodrigues1993; @Colditz1994; @Roy2014] but to have variable protection in adults ranging from 0-80%.[Mangtani2014; @Zwerling2011] This variation in protection is thought to be linked to the equator with the vaccine becoming increasing effective at higher, and lower latitudes. In England, an MRC trial in the 1950's found that the BCG vaccine was highly effective.[@Hart1972] There is little evidence to suggest that this has changed in the UK born population.

@@

\begin{equation}
\frac{dC}{dt} =  J_{d}C + J_{p}C
  (\#eq:split-models)
\end{equation}

The disease model ($J_{d}C$) is then defined as,

@@

The demographic and vaccination model ($J_{p}C$) is then defined as ($A = H,\ L,\ P,\ E,\ T_P,\ T_E$),

@@

This functional form was chosen as it is flexible enough to represent exponential growth, log constrained growth and near linear growth depending on the choice of $\iota_{\text{scale}}$. This allows the scale up of non-UK born cases to be fitted to the available data during the model fitting stage (Chapter \@ref(model-fitting)). To incorperate the uncertainty in the number of observed non-UK born cases a normal distribution was used, with the standard deviation and mean determined using parameters from the observation model (Chapter \@ref(model-fitting)).

@@

Vynnycky et al. found that the effective contact rate for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since the early 1980s and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is normally distributed with a mean of 1 and a standard deviation of 0.5. For the historic effective contact rate I have assumed a uniform distribution with a lower bound of the current effective contact rate and an upper bound of 20. I have also assumed that the historic contact rate declines over time, starting in 1935, reducing to the current effective contact rate in 1980. The speed of this decay is set by defining a half-life ($C^{\text{half-life}}_{\text{eff}}$). The prior for the half-life is assumed to be normally distributed with a mean of 5 years and a standard deviation of 5 (truncated to be greater than 0). This prior is based on the observed trend in notifications. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation explored for young adults (15-29; $\beta_{\text{young adult}}$) as a scenario. The prior for this modifier was assumed to be uniform, bounded by 0 and 10. The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

In the model presented here unstratified social (nonphysical) contacts are used to generate an age-stratified contact matrix. There are several reasons for this. Firstly, stratifying by home, school, work, transport or leisure contacts, whilst initially appealing as doing so may lead to insights as to the nature of the type of contacts required for TB transmission, may lead to over-fitting without a strong apriori hypothesis. In high and medium burden countries it has been shown that within household transmission is not a major driver of overall transmission.[@Mathema2018] Until recently it has been thought that household transmission plays a more dominated role in low burden settings, such as England, which would indicate that home contacts should be considered. However, it has recently been found that 7.7% (1849/24,060) of cases in England between 2010 and 2012 lived in a household with another case.[@Lalor2017] The same study estimated that overall only 3.9% of cases were due to recent household transmission, and there was no evidence that cases within households were more likely to transmit within the household than outside of it. There is little evidence to suggest that school, work, transport or leisure contacts are more likely to transmit TB in England than any other contact. The choice of contact type is disease dependent; for TB it is likely that closer contacts result in a greater likelihood of transmission.[@Mathema2018] Unfortunately the physical contacts recorded in the POLYMOD survey represent a poor proxy to closeness of contacts as physical contact can be a little as a handshake and because TB is a respiratory disease physical contact is not required. For this reason physical contacts have not been further evaluated. Instead, the uncertainty in age-dependent transmission rates has been explored by allowing for scenarios in which the transmission probability varies across age groups.

@@

There is strong evidence that the overall effectiveness of the BCG vaccine reduces over time.[@Abubakar2013; @Mangtani2017] For this reason the effectiveness of the BCG vaccination overall ($\alpha^T_a$) has been stratified by the time since vaccination (by 5 year age groups). This step-wise approach has been chosen as the majority of studies report estimates for these groups and the precise functional form of the reduction in protection is unknown. For 0-4, and 5-9, years since vaccination estimates of the effectiveness of the BCG vaccine were extracted from the MRC trial.[@Hart1972] Using published data, Poisson regression was used to estimate Rate Ratios and 95% confidence intervals. For 10-29 years after vaccination Rate Ratio estimates from a more recent case control cohort study in the UK born vaccinated at school-age have been used.[@Mangtani2017] Table \@ref(tab:bcg-eff-est) details the estimated effectiveness for each five yearly band after initial vaccination from these data sources. I have assumed that the BCG vaccine is equally effective regardless of when the vaccine is given as there is no evidence that protection reduces when given to older age groups in England.

@@

In this chapter, I have outlined the requirements for a dynamic transmission model of TB in order for it to be able to answer policy relevant questions relating to BCG vaccination in England. I then outlined, and gave the equations for, a semi-stochastic model that met these requirements and made best use of the data available. I defined prior distributions for each model parameter and initialisation conditions. I then detailed the data sources used for parameterisation, approximations required to make best use of the available data, and the scenario analyses needed to explore model, parameterisation and initialisation assumptions.

@@

The semi-stochastic transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future. Both of these scenarios are explored in the final chapter of this thesis.

@@

- This chapter presents a semi-stochastic transmission dynamic model of TB transmission and BCG vaccination. The model includes; age structure, pulmonary and extra-pulmonary TB, re-infection and re-activation, serial latency, TB treatment, treatment failure, TB mortality, non-UK born cases and details of the historic TB endemic. Code for this model is available online.^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi]

>>>>

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have indirect impacts, via onwards transmission, not captured in this estimate. Both of these limitations can be overcome using a dynamic transmission model (see Chapter \@ref(introduction)). This explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes, via model simulation.

This chapter presents the development and parameterisation of a dynamic model of TB transmission, incorporating BCG vaccination, in the UK born population of England. The key features of TB transmission, and BCG vaccination, are discussed with details of pertinent TB models given. An appropriate model structure for answering the study question is then outlined, along with a justification of the choices made and details of required sensitivity analyses. The model structure is then defined mathematically and parameterised using literature sources as well as data from the Enhanced TB Surveillance System (ETS), Labour Force Survey (LFS) and Office for National Statistics (ONS) (see Chapter \@ref(data)). The assumptions made during model building and parameterisation are highlighted in preparation for evaluation during model fitting (Chapter \@ref(model-fitting)).

@@

1. **Latency** - after an initial infection 5-10% of individuals develop symptomatic TB within 1-2 years. The majority of individuals enter a latent state in which they passively carry *TB mycobacteria* but are not symptomatic. Reactivation of the bacilli can then occur many years later due to a lose of immune control.[@Gideon2011a] Simplistically latent TB may be modeled with a single latent compartment[@Brooks-Pollock2010a], more commonly an additional transition rate between the susceptible and active disease states is added.[@Menzies2018] This represents rapid progression to active disease, and slower progression via a low risk latent stage. Both of these model structures have been shown to not fit activation data well.[@Menzies2018; @Ragonnet2017] More complex structures that are commonly used incorporate either parallel or serial latency (Figure \@ref(fig:latency-flow-diag)). Both of these structures incorporate both slow and fast latent periods and have been shown to produce identical activation dynamics.[@Ragonnet2017] This is unfortunate as they represent two disparate biological mechanisms, with the serial assumption representing decreasing risk over time for individuals and the parallel assumption suggesting that a subset of individuals are at a greater risk of developing active TB disease. For models that seek to investigate interventions targeted at latent cases this structural uncertainty is problematic. However, as BCG vaccination occurs prior to infection both structures will produce comparable results for study questions evaluating this intervention. The model presented here uses a serial latent structure. This is commonly used in the literature; simplifies modelling other aspects of TB; and has a plausible biological underpinning.[@Ragonnet2017]

@@

1. **Protection from active disease** - the BCG vaccine has been shown to primarily protect against the progression from latent to active TB disease (Chapter \@ref(background)). It has been shown to be highly protective in children[@Rodrigues1993; @Colditz1994; @Roy2014] but to have variable protection in adults ranging from 0-80%.[Mangtani2014; @Zwerling2011] This variation in protection is thought to be linked to the equator with the vaccine becoming increasing effective at higher, and lower latitudes. In England, an MRC trial in the 1950s found that the BCG vaccine was highly effective.[@Hart1972] There is little evidence to suggest that this has changed in the UK born population.

@@

\begin{equation}
\frac{dC}{dt} =  J_{d} + J_{p}
  (\#eq:split-models)
\end{equation}

The disease model ($J_{d}$) is then defined as,

@@

The demographic and vaccination model ($J_{p}$) is then defined as ($A = H,\ L,\ P,\ E,\ T_P,\ T_E$),

@@

This functional form was chosen as it is flexible enough to represent exponential growth, bounded growth and near linear growth depending on the choice of $\iota_{\text{scale}}$. This allows the scale up of non-UK born cases to be fitted to the available data during the model fitting stage (Chapter \@ref(model-fitting)). To incorperate the uncertainty in the number of observed non-UK born cases a normal distribution was used, with the standard deviation and mean determined using parameters from the observation model (Chapter \@ref(model-fitting)).

@@

Vynnycky et al. found that the effective contact rate for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since the early 1980s and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is normally distributed with a mean of 1 and a standard deviation of 0.5. For the historic effective contact rate I have assumed a uniform distribution with a lower bound of the current effective contact rate and an upper bound of 20. I have also assumed that the historic contact rate declines over time, starting in 1935, reducing to the current effective contact rate in 1980. The speed of this decay is set by defining a decay parameter ($C^{\text{half-life}}_{\text{eff}}$). The prior for the decay parameter is assumed to be normally distributed with a mean of 5 years and a standard deviation of 5 (truncated to be greater than 0). This prior is based on the observed trend in notifications. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation explored for young adults (15-29; $\beta_{\text{young adult}}$) as a scenario. The prior for this modifier was assumed to be uniform, bounded by 0 and 10. The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

In the model presented here unstratified social (nonphysical) contacts are used to generate an age-stratified contact matrix. There are several reasons for this. Firstly, stratifying by home, school, work, transport or leisure contacts, whilst initially appealing as doing so may lead to insights as to the nature of the type of contacts required for TB transmission, may lead to over-fitting without a strong a priori hypothesis. In high and medium burden countries it has been shown that within household transmission is not a major driver of overall transmission.[@Mathema2018] Until recently it has been thought that household transmission plays a more dominated role in low burden settings, such as England, which would indicate that home contacts should be considered. However, it has recently been found that 7.7% (1849/24,060) of cases in England between 2010 and 2012 lived in a household with another case.[@Lalor2017] The same study estimated that overall only 3.9% of cases were due to recent household transmission, and there was no evidence that cases within households were more likely to transmit within the household than outside of it. There is little evidence to suggest that school, work, transport or leisure contacts are more likely to transmit TB in England than any other contact. The choice of contact type is disease dependent; for TB it is likely that closer contacts result in a greater likelihood of transmission.[@Mathema2018] Unfortunately the physical contacts recorded in the POLYMOD survey represent a poor proxy to closeness of contacts as physical contact can be a little as a handshake and because TB is a respiratory disease physical contact is not required. For this reason physical contacts have not been further evaluated. Instead, the uncertainty in age-dependent transmission rates has been explored by allowing for scenarios in which the transmission probability varies across age groups.

@@

There is strong evidence that the overall effectiveness of the BCG vaccine reduces over time.[@Abubakar2013; @Mangtani2017] For this reason the effectiveness of the BCG vaccination overall ($\alpha^T_a$) has been stratified by the time since vaccination (by 5 year age groups). This step-wise approach has been chosen as the majority of studies report estimates for these groups and the precise functional form of the reduction in protection is unknown. For 0-4, and 5-9, years since vaccination estimates of the effectiveness of the BCG vaccine were extracted from the MRC trial.[@Hart1972] Using published data, Poisson regression was used to estimate rate ratios and 95% confidence intervals. For 10-29 years after vaccination rate ratio estimates from a more recent case control cohort study in the UK born vaccinated at school-age have been used.[@Mangtani2017] Table \@ref(tab:bcg-eff-est) details the estimated effectiveness for each five yearly band after initial vaccination from these data sources. I have assumed that the BCG vaccine is equally effective regardless of the age at vaccination as there is no evidence that protection reduces when given to older age groups in England.

@@

In this chapter, I have outlined the requirements for a dynamic transmission model of TB in order for it to be able to answer policy relevant questions relating to BCG vaccination in England. I then outlined, and gave the equations for, a model that met these requirements and made use of the data available. I defined prior distributions for each model parameter and initialisation conditions. I then detailed the data sources used for parameterisation, approximations required to make best use of the available data, and the scenario analyses needed to explore model, parameterisation and initialisation assumptions.

@@

The transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future. Both of these scenarios are explored in the final chapter of this thesis.

@@

- This chapter presents a transmission dynamic model of TB transmission and BCG vaccination. The model includes; age structure, pulmonary and extra-pulmonary TB, re-infection and re-activation, serial latency, TB treatment, treatment failure, TB mortality, non-UK born cases and details of the historic TB endemic. Code for this model is available online.^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi]
